\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-23 chapters/wcble.tex
Literature wcble

Visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli.
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform.
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological and clinical properties, that can be confounding factors in
the development of \acp{bci}.\todo{cite}
Latency is often assessed as the latency of \ac{erp} components in a subject
average or grand average.
Latencies of non-averaged, single-trial \ac{erp} waveforms could capture even more
relevant information, yet they are very hard to measure due to low \ac{snr} of
single trial \acp{erp}.
Within session variability of \ac{erp} latency, or \ac{erp} latency
\emph{jitter} has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting with patients in the intended target population.
\textcite{Zisk2021} showed that patients with \ac{als} have significantly
increased ERP jitter.
\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
performance, and disproportionally so when the user is not directly gazing at
intended targets, such as is the case for for gaze-impaired patients.

@@

\subsubsection{Benefits of latency estimation}
A common approach to counteract the smearing effect is to correct single-trial
\ac{erp}s for jitter before averaging.\todo{reference}
This can be done by estimating the latency of each
single-trial \ac{erp}, and aligning them before averaging.
Latency estimation has multiple advantages:
\begin{enumerate*}[label={(\arabic*)}]
	\item Latency-based \ac{erp} alignment can help reveal the true shape and
	properties like peak amplitude and latency of \ac{erp}s across
	experimental conditions. These are variables of interest for
	testing electrophysiological an behavioral hypotheses.
	Additionally, latency estimation allows for the extraction of
	other descriptors from a set of single-trial \ac{erp}s, like
	amplidute or latency variance~\cite{Hultsch2004}, which can
	also be of interest as studied variables.
  For instance, in the work of \textcite{Saville2014}, the inter-trial latency
  variability itself is determined as a correlate of working memory performance.
	\item Counteracting the smearing effect will boost the signal-to-noise
	ratio of averaged \ac{erp}s by increasing the amplitude of the
	\ac{erp} signal. This in turn can reveal smaller or more jittered \ac{erp}
  components which would otherwise obscured or blurred due to desynchronization.
	These deblurred templates can then be used for analysis of the \ac{erp}
	waveform, or as templates of the signal of interest in
	BCIs~\cite{arico2014influence}.
	%this enables decoders to take into account the relatively low amplitude
	%earlier components, leading to applications in settings where
	%information from other components than the P300 need to be exploited,
	%like covert visual attention paradigms~\cite{arico2014influence,
	%hardiansyah2020single}.\todo[inline]{rework this to latency features
	%because smaller visual components might not really be revealed
	%since they're already time-locked}
	\item Finally, the introduction of latency estimation and alignment generaly
	improves BCI classification performance. Single-trial latencies can be
	used as features to improve classification~\cite{hardiansyah2020single}.
	Furthermore, compensating for inter-trial differences, especially when
	extended to inter-subject differences, can aid in transfer learning or
	generalization across subjects and protocols\cite{Iturrate2014}.
\end{enumerate*}
\todo{Read, summarize and cite BCI	references from bibliography}

@@

The most straight-forward method to estimate single-trial latency
peak-picking. Here, the latency of an \ac{erp} is determined as the
time-point of it's maximum (or minimum) amplitude with relaion to the stimulus
onset. While straight-forward, this method
does not perform well in low or negative SNR conditions, unless combined with
filtering. Filtering can suppress the noise contaminating EEG trials, lowering
the risk of picking a noise peak instead of a true peak of the \ac{erp}. Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{Carlton1980, McCarthy1981,
	Karjalainen1999, Nishida1999,  Sparacino2002, Georgiadis2005,
	D’Avanzo2011} and in the time-frequency
domain~\cite{Quiroga2003, wang2007single,Hu2010,arico2014influence}. Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using PCA~\cite{Ouyang2017}, ICA~\cite{townsend1999analyzing, jung2001analysis,
	Milne2011}, or as a spatial LCMV beamformer~\cite{TREDER2016279}.

A generally more preferred approach than
peak-picking is template matching. Woody's algorithm~\cite{woody1967characterization} is a simple and performant latency
estimation method that is still being used and is considered amongst the more
performant techniques~\cite{Ouyang2017}. It starts with the average \ac{erp} as a
rough estimation of the aligned template \ac{erp}, and iteratively refines this
template by determining at each step the time point of maximum
cross-correlation of this template with all single-trial \ac{erp}s and aligning them
on these time points to form the next template.
Similarily, cross-correlation template matching can be combined with filtering.
\cite{souloumiac2013improved} uses XDAWN to construct spatial filter before
template matching the spatially filtered signal. \cite{Iturrate2014} uses a
spatiotemporal filter. Instead of aligning the
latencies to obtain the template for the next iteration, weighted averaging can
also be used~\cite{gasser1983selavco}.

Despite good performance, the iterative scheme of Woody's algorithm has a risk of converging to a local
optimum. To avoid this, a fitness function can be defined to jointly optimize
the set of single-trial \ac{erp} latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.

Methods adapted to handle multiple conditions, useful in BCI setting
\begin{itemize}
	\item \cite{Kim2020}
\end{itemize}

\subsection{Multi-component approaches}
\label{sec:wcble/literature/multi-comp}
The abovementioned methods do not take into account the fact that  \ac{erp}s are multi-component signals. Multiple components can be time-locked to different events or neurophysiological proccess and hence are not necessarily
time-locked to each other. They each can have different latency
variabilities. The presence of more than one component can hamper the
performance of some algorithms, and it often only possible to extract only the
latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
to one component and averaging, this will introduce blurring in the other
components~\cite{Ouyang2020}.

Most of the algorithms above can be adapted to work on multiple \ac{erp} components
by carefully selecting peaks or pre-determined regions of interest.
\cite{hardiansyah2020single} uses an SNR boosting method, applied to a specific
time-window for each \ac{erp} component to extract a latency per component.
Methods based on spatial decomposition filters like PCA, ICA or XDAWN can be leveraged
to utilize multiple of their components as filters, resulting in multiple time
series where peack picking or Woody's algorithm can be applied. In general,
however, these algorithms lack an integrated approach to deal with multiple
\ac{erp} components. A clear evaluation is lacking of the performance of these
methods in
multi-component settings is lacking.

Some algorithms separate components into stimulus and response locked component clusters based on their
latency distributions~\cite{jung2001analysis, Takeda2008,
	Zhang1998, Yin2009}. These algorithms do, however, require a response
button trigger signal as input to determine the response time, which is
not always present or applicable in an \ac{erp} experiment, especially in
the case of BCIs.
The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,
	Wang2015, Ouyang2016, Ouyang2020} is able to do this without entirely depending on
the response time, and can separate stimulus-locked, response-locked,
and multiple central component clusters. Others, like spatial filtering methods that use
decomposition methods to determine the spatial filter can be be modified to separate
components.~\cite{Ouyang2017}.

To categorize:\todo{categorize and analyze these references}
\begin{itemize}
	\item \cite{Dimitriadis2018}
	\item \cite{Thornton2008}
	\item Maximum likelihood~\cite{tuan1987variable, Moecks1988,Puce1994} Maximum likelihood
	\item \cite{Gratton1989} (no access) Maximum likelihood?
\end{itemize}

\todo{Explain beamforming}

Finally, the smearing problem can also be overcome by algorithms that are not
based on \ac{erp} component latency estimation, but which use time warping-based methods
to enhance the shape of the averaged template \ac{erp}, like Dynamic Time
Warping~\cite{486255, wang2001warp, Zoumpoulaki2015}, Correlation-Optimized
Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{flotho2021fast}. Since these methods work
fundamentally different from the latency estimation methods, and cannot
directly be used to extract latency features, we will not further consider them
in this work.

\todo{check overlap with:

The most straightforward latency estimation method is \textit{peak picking}.
Peak picking determines the latency of an \ac{erp} as the time point of its maximum
(or minimum for negative components) amplitude relative to stimulus
onset.
While easy to implement, this method does not perform well in low or negative
SNR conditions because of the risk of picking a noise-related peak, unless
combined with strong filtering in the space, time or frequency
domains~\cite{Avanzo2011,Arico2014,TREDER2016279,Ouyang2017}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Due to the risk of picking noise peaks, template matching is generally
preferred.
Woody's algorithm~\cite{woody1967characterization} is a simple and performant
latency estimation method that is considered to be among
the most performant ones~\cite{Ouyang2017}.
Woody's scheme starts with a template that approximates the expected response,
e.g. the average target \ac{erp}, and iteratively enhances the SNR of this template by determining at
each iteration the time point of maximum cross-correlation of the template
with all single-trial \ac{erp}s, and aligning them based on these latencies to form
the template for the next iteration.
The final, enhanced template can then be cross-correlated again to retrieve an
accurate estimation of single-trial latencies.
Similar to peak-picking, cross-correlation template matching can be combined
with a spatial filter such as XDAWN~\cite{souloumiac2013improved} or a
spatiotemporal filter \cite{Iturrate2014}.
More recent work achieves good performance in latency estimation using residue
iteration~\cite{Ouyang2017}, genetic
algorithms~\cite{Pelo2018}, a hidden process model~\cite{Kim2020} or
graph-based network analysis~\cite{Dimitriadis2018}.

When multiple clusters of \ac{erp} components are not time-locked to
each other, aligning to the latency of one cluster might introduce more jitter
in other component clusters, further decreasing the SNR~\cite{Ouyang2020}.

Finally, algorithms that use time
warping-based methods instead of explicit \ac{erp} component latency estimation can
also overcome the smearing problem.
Examples that have been applied to \ac{erp}s are Dynamic Time
Warping~\cite{Casarotto2005,486255,wang2001warp,Zoumpoulaki2015},
Correlation-Optimized Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{Flotho2021}. Since these methods are fundamentally
different from latency estimation methods,
we will not further consider them in this work.

}

@@

@@

\section{Results}

@@

\section{Conclusion}

>>>>

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli.
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform.
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological and clinical properties, that can be confounding factors in
the development of \acp{bci}.\todo{cite}
Latency is often assessed as the latency of \ac{erp} components in a subject
average or grand average.
Latencies of non-averaged, single-trial \ac{erp} waveforms could capture even more
relevant information, yet they are very hard to measure due to low \ac{snr} of
single trial \acp{erp}.
Within session variability of \ac{erp} latency, or \ac{erp} latency
\emph{jitter} has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting with patients in the intended target population.
\textcite{Zisk2021} showed that patients with \ac{als} have significantly
increased ERP jitter.
\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
performance, and disproportionally so when the user is not directly gazing at
intended targets, such as is the case for for gaze-impaired patients.

@@

\subsubsection{Benefits of latency estimation}
A common approach to counteract the smearing effect is to correct single-trial
\ac{erp}s for jitter before averaging.\todo{reference}
This can be done by estimating the latency of each
single-trial \ac{erp}, and aligning them before averaging.
Latency estimation has multiple advantages:
\begin{enumerate*}[label={(\arabic*)}]
	\item Latency-based \ac{erp} alignment can help reveal the true shape and
	properties like peak amplitude and latency of \ac{erp}s across
	experimental conditions. These are variables of interest for
	testing electrophysiological an behavioral hypotheses.
	Additionally, latency estimation allows for the extraction of
	other descriptors from a set of single-trial \ac{erp}s, like
	amplidute or latency variance~\cite{Hultsch2004}, which can
	also be of interest as studied variables.
  For instance, in the work of \textcite{Saville2014}, the inter-trial latency
  variability itself is determined as a correlate of working memory performance.
	\item Counteracting the smearing effect will boost the signal-to-noise
	ratio of averaged \ac{erp}s by increasing the amplitude of the
	\ac{erp} signal. This in turn can reveal smaller or more jittered \ac{erp}
  components which would otherwise obscured or blurred due to desynchronization.
    These \emph{deblurred} templates can then be used for analysis of the \ac{erp}
	waveform, or as templates of the signal of interest in
	BCIs~\cite{Arico2014}.
	%this enables decoders to take into account the relatively low amplitude
	%earlier components, leading to applications in settings where
	%information from other components than the P300 need to be exploited,
	%like covert visual attention paradigms~\cite{arico2014influence,
	%hardiansyah2020single}.\todo[inline]{rework this to latency features
	%because smaller visual components might not really be revealed
	%since they're already time-locked}
	\item Finally, the introduction of latency estimation and alignment generally
	improves BCI classification performance.
  Single-trial latencies can be	used as features to improve
  classification~\cite{Hardiansyah2020}.
	Furthermore, compensating for inter-trial differences, especially when
	extended to inter-session or inter-subject differences, can aid in transfer learning or
	generalization across subjects and protocols\cite{Iturrate2014}.
\end{enumerate*}

@@

The most straight-forward method to estimate single-trial latency is
\emph{peak picking}
Here, the latency of an \ac{erp} is determined as the
time-point of it's maximum (or minimum) amplitude relative to stimulus
onset.
While straight-forward, this method does not perform well in low or negative
\ac{snr} conditions, unless combined with filtering.
Filtering can suppress the noise contaminating \ac{eeg} trials, lowering
the risk of picking a noise peak instead of a true peak of the \ac{erp}.
Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{
  Carlton1980, McCarthy1981, Karjalainen1999, Nishida1999,  Sparacino2002,
  Georgiadis2005,	D’Avanzo2011}
and in the time-frequency domain~\cite{
  Quiroga2003, Wang2007,Hu2010}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using \ac{pca}~\cite{Ouyang2017}, ICA~\cite{Townsend1999, Jung2001,
	Milne2011}, or as a spatial LCMV beamformer~\cite{Treder2016}.

Template matching is generally preferred to peak picking.
The algorithm proposed by \textcite{Woody1967} in 1976 is a simple and elegant
latency estimation method that is still being used and is even
considered amongst the more performant techniques for \acp{erp}~\cite{Ouyang2017}.
It starts with an average \ac{erp} as a rough estimation of the aligned template
\ac{erp}, and iteratively refines this template by determining at each step the time point of maximum
cross-correlation of this template with all single-trial \acp{erp} and aligning them
on these time points to form the next template.
Similarily, cross-correlation template matching can be combined with filtering.
\cite{Souloumiac2013} uses XDAWN to construct spatial filter before
template matching the spatially filtered signal. \cite{Iturrate2014} uses a
spatiotemporal filter. Instead of aligning the
latencies to obtain the template for the next iteration, weighted averaging can
also be used~\cite{Gasser1983}.
\todo{cite Thornton2008 hierarchical woody?}

Despite good performance, the iterative scheme of Woody's algorithm has a risk of converging to a local
optimum.
To avoid this, a fitness function can be defined to jointly optimize
the set of single-trial \ac{erp} latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.
Alternative methods are based on graph
optimization~\cite{Dimitriadis2018}, hidden process models~\cite{Kim2020} or by fitting an \ac{erp} model with
maximum likelihood estimation~\cite{Gratton1989, Tuan1987, Moecks1988,Puce1994}.

\subsection{Multi-component approaches}
\label{sec:wcble/literature/multi-comp}
In the previous section, we have abstracted away the fact that \ac{erp}s are
multi-component signals, which is not taking into account by the aforementioned
methods.
Multiple components can be time-locked to different events or neurophysiological
proccess and therefore are not necessarily time-locked to each other. They
each can have different latency variabilities.
The presence of more than one component can hamper the
performance of some latency estimation algorithms, and it often only possible to extract only the
latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
to one component and averaging, this will introduce blurring in the other
components~\cite{Ouyang2020}.

Most of the algorithms above can be adapted to work on multiple \ac{erp} components
by carefully selecting peaks or pre-determined regions of interest.
\cite{Hardiansyah2020} uses an \ac{snr} boosting method, applied to a specific
time-window for each \ac{erp} component to extract a latency per component.
Methods based on spatial decomposition filters like \ac{pca}, ICA or XDAWN can be leveraged
to utilize multiple of their components as filters, resulting in multiple time
series where peack picking or Woody's algorithm can be applied. In general,
however, these algorithms lack an integrated approach to deal with multiple
\ac{erp} components. A clear evaluation is lacking of the performance of these
methods in
multi-component settings is lacking.

Some algorithms separate components into stimulus and response locked component clusters based on their
latency distributions~\cite{Jung2001, Takeda2008,	Zhang1998, Yin2009}.
These algorithms do, however, require a response
button trigger signal as input to determine the response time, which is
not always present or applicable in an \ac{erp} experiment, especially in
the case of BCIs.
The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,	Wang2015, Ouyang2016, Ouyang2020}
is able to do this without entirely relying on the response time, and can
separate stimulus-locked, response-locked, and multiple central component clusters.
Others, like spatial filtering methods that use decomposition methods to
determine the spatial filter can be be modified to separate
components~\cite{Ouyang2017}.

Finally, the smearing problem can also be overcome by algorithms that are not
based on \ac{erp} component latency estimation, but which use time warping-based methods
to enhance the shape of the averaged template \ac{erp}, like Dynamic Time
Warping~\cite{486255, wang2001warp, Zoumpoulaki2015}, Correlation-Optimized
Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{flotho2021fast}.
These methods work
fundamentally different from the latency estimation methods, and cannot
directly be used to extract latency features.

@@

\todo{cite Kim2020 used to analyze the difference between target and
non-target}

@@

\section{Results}
simulation study
refer to later chapter for use on real erp data

@@

\section{Conclusion}
Current work is focusing on the extension to multi-component algorithm, in a
similar fashion to ouyang
